---
title: "Two-stage cluster sampling estimation: variance"
author: "Stat 260, St. Clair"
subtitle: "Week 7 (5.3)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, comment=NULL,tidy.opts=list(width.cutoff=60),tidy=FALSE,eval=TRUE,eval=TRUE, warning=FALSE, message  = FALSE, fig.width = 11, fig.height = 5)
library(dplyr)
library(ggplot2)
library(survey)
schools <- read.csv("http://math.carleton.edu/kstclair/data/california_cluster.csv")
schools$N <- 757  # N
schools$n <- 40   # n 
schools <- schools %>%
    group_by(dnum) %>% # group by cluster (district number)
    mutate(m_i = n())  #  m_i = sample size per cluster
summary(schools$m_i)
schools$wts <- (757/40)*schools$district_size/schools$m_i
schools_design <- svydesign(id= ~dnum + snum, 
                            fpc= ~N + district_size, 
                            weights = ~wts, 
                            data=schools)
```




## Population Total: Two-Stage Cluster 

- **Parameter**: $t = \sum_{i=1}^N \sum_{j=1}^{M_i} y_{ij}=  \sum_{i=1}^N t_i$

- **Unbiased Estimator:** 
$$\hat{t}_{unb} =  \sum_{i =1}^{n}  \dfrac{N}{n} M_{i} \bar{y}_{i} = \sum_{i =1}^{n} \dfrac{N}{n} \hat{t}_{i}$$
where $\bar{t}$ is the sample mean **total response** per cluster.

- **Standard error:**
$$SE(\hat{t}_{unb}) =  \sqrt{N^{2}\left(1- \dfrac{n}{N} \right) \dfrac{s^{2}_{t}}{n} + \dfrac{N}{n} \sum_{i=1}^n  \left(1- \dfrac{m_{i}}{M_{i}} \right) M_{i}^{2}\dfrac{s^{2}_{i}}{m_{i}} }$$

---

## Variance: between + within cluster variation

**1. Between cluster variation:** $N^{2}\left(1- \dfrac{n}{N} \right) \dfrac{s^{2}_{t}}{n}$

  - just the **one-stage** variance 

<br>
<br>
<br>

**2. Within cluster variation:**  $\dfrac{N}{n} \sum_{i=1}^n  \left(1- \dfrac{m_{i}}{M_{i}} \right) M_{i}^{2}\dfrac{s^{2}_{i}}{m_{i}}$

  - added because we don't sample *all* units within a cluster

<br>
<br>

- Same explanation of the *ratio* estimator $SE(\hat{\bar{y}}_r)$

---

## Comparing cluster sampling and SRS

- **One-stage** *less* precise than a SRS when clusters are *homogeneous*

  - Same holds true for **two-stage**
  
  - Generally, $SE_{two-stage} \geq  SE_{one-stage} > SE_{SRS}$

---

## Two-stage cluster sampling: Assessing homogeneity

- Visually: plot cluster id vs $y_{ij}$ (boxplot, scatterplot)

- Adjusted R-squared: 
$$
\hat{R}^{2}_{a} = 1-\dfrac{\widehat{MSW}}{\hat{S}^{2}}
$$
- $MSW$ is estimated from **sample ANOVA** values  $msw$

- **Cluster/sample sizes equal:**  $S^2$ is estimated from **sample ANOVA** values $msb$ and $msw$:
$$
\hat{S}^2 = \dfrac{\dfrac{M}{m}(N-1)msb + \left(\dfrac{m-1}{m}NM + \dfrac{M}{m}-1\right)msw}{NM-1} 
$$

- **Cluster/sample sizes NOT equal:** best "by hand" (but biased) approximation is the sample variance of all responses $s^2$


---

## Example: California API scores by district

```{r}
schools_by_district <- schools %>% 
  group_by(dnum) %>%  # group by cluster (district number)
  summarize( s_i = sd(api00),    # SD by cluster
            m_i = n(), # sample size per cluster
            M_i = first(district_size) ,
            samp_frac = m_i/M_i)
summary(schools_by_district)
```


---

## Example: California API scores by district

```{r, echo=TRUE}
ggplot(schools, aes(x=dname, y = api00)) + 
  geom_point() + 
  geom_vline(aes(xintercept=dname), linetype= "dotted") + 
  coord_flip() + 
  labs(x="District (cluster)")
```



---

## Example: California API scores by district

- Cluster (district) sizes in the population are *not* similar. 

  - quick approximation gives $R^2_a  \approx 0.86$

```{r}
api_lm <- lm(api00 ~ dname, data=schools)
anova(api_lm)
1 - 2567/var(schools$api00)   # r^2_a =  1=msw/s^2
summary(api_lm)$adj.r.squared  # r^2_a 
```


---

## Example: California growth by district

```{r, echo=TRUE}
ggplot(schools, aes(x=dname, y = growth)) + 
  geom_point() + 
  geom_vline(aes(xintercept=dname), linetype= "dotted") + 
  coord_flip() + 
  labs(x="District (cluster)")
```


---

## Example: California growth by district

- quick approximation gives $R^2_a  \approx 0.24$


```{r}
growth_lm <- lm(growth ~ dname, data=schools)
summary(growth_lm)$adj.r.squared
```


---

## Example: California growth by district

- `growth` is more heterogeneous by district so its SE is more similar to a SRS than `api00`

```{r}
# design from estimation slides
svymean(~api00 + growth, schools_design, deff = TRUE)
```

---

## Variance: between + within cluster variation

$$Var(\hat{t}_{unb}) =  N^{2}\left(1- \dfrac{n}{N} \right) \dfrac{S^{2}_{t}}{n} + \dfrac{N}{n} \sum_{i=1}^N  \left(1- \dfrac{m_{i}}{M_{i}} \right) M_{i}^{2}\dfrac{S^{2}_{i}}{m_{i}}$$
<br>

**Sketch of Proof**
Uses Appendix Section A.4 Rule 5:
$$
Var(Y) = Var_y(E_x(Y \mid X)) + E_y(Var_x(Y \mid X))
$$



---

## A. between  cluster variation

Fixed clusters (stage 1): $i \in \mathcal{S}$ are fixed
$$E_2(\hat{t} \mid \textrm{stage 1 fixed})$$



---

## A. between  cluster variation

How does $E_2$ vary across stage 1 samples?
$$Var_1(E_2(\hat{t} \mid \textrm{stage 1 fixed}))$$



---

## B. within  cluster variation

Fixed clusters (stage 1): $i \in \mathcal{S}$ are fixed
$$Var_2(\hat{t} \mid \textrm{stage 1 fixed})$$



---

## B. within  cluster variation

What is the expected $V_2$ across stage 1 samples?
$$E_1(Var_2(\hat{t} \mid \textrm{stage 1 fixed}))$$
